---
title: "The exactmed function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The exactmed function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  prompt = TRUE,
  comment = " "
)
```

## Introduction

This document aims to illustrate the usage of the function `exactmed()` and its behavior via additional examples.

 We recall that `exactmed()` computes natural direct and indirect effects, and controlled direct effects for a binary outcome and a binary mediator. The user can specify the reference levels of the outcome and mediator variables using the input parameters `hvalue_m` and `hvalue_y`, respectively (see the function help). Controlled direct effects are obtained for both possible mediator values ($m=0$ and $m=1$). Natural and controlled effects can either be unadjusted or adjusted for covariates (that is, conditional effects). Adjusted effect estimates are obtained for covariates fixed at their sample-specific mean values (default) or at specific values (user-provided). Also, by default, `exactmed()` incorporates a mediator-exposure interaction term in the outcome model, which  can be removed by setting `interaction=FALSE`. Concerning interval estimates, `exactmed()`  generates, by default, $95\%$ confidence intervals obtained by the delta method. Alternatively, bootstrap confidence intervals, instead of delta method confidence intervals, can be obtained by specifying `boot=TRUE` in the function call. In this case, 1000 bootstrap datasets are generated by default.
 

 
## The dataset
 
 In all the examples presented below we use the dataset `datamed`, available after loading the **ExactMed** package. Some of the features of this dataset can be found in its corresponding help file (`help(datamed)`). We recall that the `exactmed()` function only works on data frames with named columns and no missing values.
 
```{r}
library(ExactMed)

head(datamed)

```
 
The following command verifies whether the dataset contains any missing values:
```{r}

as.logical(sum(is.na(datamed)))

```
 
 
## Examples 

### Basic examples
 
Suppose that one wishes to obtain mediation effects estimates for a change in exposure from $0$ to $1$, assuming there is no exposure-mediator interaction and using the delta method to construct $95\%$ confidence intervals.

In this case, a valid call to `exactmed()` would be:
 

```{r}

results1 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', 
  a1 = 1, a0 = 0, interaction = FALSE
  )  

results1

```
 
 
Mediation effects estimates adjusted for covariates are obtained through the use of the character vectors `m_cov` and `y_cov`, which contain the names of the covariates to be adjusted for in the mediator and outcome models, respectively. The following call to `exactmed()` incorporates covariates `C1` and `C2` in both the mediator and outcome models:

```{r}

results2 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0,  
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  interaction = FALSE
  )

results2

```


`exactmed()` also allows for the specification of two different sets of covariates in the mediator and outcome models. For example, the following specification of `m_cov` and `y_cov` means that the mediator model is adjusted for `C1` and `C2`, while the outcome model is adjusted for `C1` only.

However, we advise against this practice unless it is known that excluded covariates are independent of the dependent variable (mediator or outcome) being modeled given the rest of covariates.  


```{r}

results3 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0,  
  m_cov = c('C1', 'C2'), y_cov = c('C1'), 
  interaction = FALSE
  )

results3

```


By default, the `adjusted` parameter is `TRUE`. If the `adjusted` parameter is set to `FALSE`, `exactmed()` ignores the values of the vectors `m_cov` and `y_cov`, and computes unadjusted (crude) effect estimates as in the first example above: 
 
```{r}

results4 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1'), 
  adjusted = FALSE, interaction = FALSE
  )

results4

```
 

To perform an adjusted mediation analysis allowing for exposure-mediator interaction (by default `TRUE`) and using bootstrap based on $100$ resamples with initial random seed $= 1991$ to construct $97\%$ confidence intervals, one should call `exactmed()` as follows:

 
```{r, results='hide'}

results5 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  boot = TRUE, nboot = 100, bootseed = 1991, confcoef = 0.97
  )

```
 
 
```{r}

results5

```


### Firth's penalization 

 
In the situation where we believe that we are facing a problem of separation or quasi-separation, Firth's penalization can be used by setting the `Firth` parameter to `TRUE` (Firth penalized mediation analysis).
If this is the case, Firth's penalization is applied to both the mediator model and the outcome model.  


The `Firth` parameter implements Firth's penalization to reduce the bias of the regression coefficients estimators under scarce or sparse data:

```{r, results='hide'}

results6 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), Firth = TRUE, 
  boot = TRUE, nboot = 100, bootseed = 1991, confcoef = 0.97
  )

```

 
```{r}

results6

```
 
 
### Stratum-specific effects

 
The following call to `exactmed()` returns mediation effects when the values of the covariates `C1` and `C2` are $0.1$ and $0.4$, respectively, assuming an exposure-mediator interaction and using the delta method to construct $95\%$ confidence intervals:
 
 
```{r}

results7 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C1 = 0.1,C2 = 0.4), y_cov_cond = c(C1 = 0.1, C2 = 0.4)
  )

results7

```
 
 
 
Common adjustment covariates in vectors `m_cov` and `y_cov` must have the same values; otherwise, the execution of the `exactmed()` function is aborted and an error message is displayed in the R console. Example:


```{r, error=TRUE, collapse=FALSE}

exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1','C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C1 = 0.3, C2 = 0.4), y_cov_cond = c(C1 = 0.1, C2 = 0.4)
 )


```


If the covariates specified in `m_cov_cond` (`y_cov_cond`) constitute some proper subset of `m_cov` (`y_cov`) then the other covariates are set to their sample-specific mean levels. Hence, the call

```{r}

results8 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C1 = 0.1), y_cov_cond = c(C1 = 0.1)
  )

```

 
 is equivalent to:
 
```{r}

 mc2 <- mean(datamed$C2)
 mc2

results9 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C1 = 0.1, C2 = mc2), y_cov_cond = c(C1 = 0.1, C2 = mc2)
  )

```
 
 
This can be checked by comparing the two outputs:
 
 
```{r}

identical(results8, results9)

```
 

With this in mind, an error is easily predicted if one makes this call:


```{r,error=TRUE, collapse=FALSE}

exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C1 = 0.1), y_cov_cond = c(C1 = 0.1, C2 = 0.4)
  )


```

 
 
### Categorical covariates

`exactmed()` also allows for categorical covariates. Covariates of this type must appear in the data frame as factor, character or logical columns. To illustrate how `exactmed()` works with categorical covariates, we replace the covariate `C1` in the dataset `datamed` by a  random factor column:

```{r}

cate <- factor(sample(c("a", "b", "c"), nrow(datamed), replace =TRUE))
datamed$C1 <- cate

```


It is possible to estimate mediation effects at specific values of categorical covariates using the input parameters `m_cov_cond` and `y_cov_cond`. Note that if the targeted covariates are both numerical and categorical, the above parameters require to be list-type vectors, instead of atomic vectors as when covariates are only numerical.
 
 Hence, if one wants to estimate mediation effects at level 'a' for `C1` and at value $0.4$ for `C2`, assuming an exposure-mediator interaction and using the delta method to construct $95\%$ confidence intervals, `exactmed()` should be called as follows:
 
 
```{r}

results10 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = list(C1 = 'a', C2 = 0.4), y_cov_cond = list(C1 = 'a', C2 = 0.4)
  )

results10

```
 

If one does not specify a value for the categorical covariate `C1`, `exactmed()` computes the effects by assigning each dummy variable (created internally by `exactmed()`) to a value equal to the proportion of observations in the corresponding category (equivalent to setting each dummy variable to its mean value): 


```{r}

results11 <- exactmed(
  data = datamed, a = 'X', m = 'M', y = 'Y', a1 = 1, a0 = 0, 
  m_cov = c('C1', 'C2'), y_cov = c('C1', 'C2'), 
  m_cov_cond = c(C2 = 0.4), y_cov_cond = c(C2 = 0.4)
  )

results11

```
 
 
 
 
 
 
 
 
 
 

